A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-09-24 main.tex
20200924 meeting notes

Furthermore, the proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ (i.e., the portion that will cause confounding) is given by
\begin{equation}
    \label{eqn:ratio}
    ||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2,
\end{equation}
which we refer to as the Confounding Proportion (CP). The CP can range from 0 (no confounding) to 1 (entirely confounded). In the extreme case where $\bZ \bgamma$ is completely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, but the residual error will be inflated.

@@

 To evaluate the performance of penalized LMM methods in the presence of these different forms of confounding complexity, we consider environmental effects which are concordant or discordant in nature in the presence of both exponential and dichotomous effect sizes, for a total of four distinct $\bgamma$ patterns. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient patterns, where the subpopulations on the x-axis are ordered by genetic similarity so that subpopulation 1 is most genetically similar to subpopulation 2, and least similar to subpopulation 10. This may also be thought of as subpopulations ordered spatially along a 1-dimensional geographic region.
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{An illustration of four distinct patterns of environmental effects and corresponding $\bgamma$ values in the presence of 10 subpopulations ordered by genetic similarity.}
    \label{fig:gamma_structures}
\end{figure}
\noindent In the case of concordant and dichotomous effects, Figure \ref{fig:gamma_structures} shows that the first five ordered subpopulations are subject to an effect size of 1, while the second five are subject to an effect size of $-$1. In the case of discordant dichotomous effects, the first five subpopulations experience an effect size with an absolute value of \anna{1}, and the second five an absolute effect size of \anna{0.5}. However, the directions of these absolute effects sizes alternate among the ordered subpopulations so that genetically similar subpopulations do not experience similar environmental effects. \anna{not quite 1 and 0.5 because of centering and standardization...how to word this so it's not confusing?}

@@

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical CP with those from the simulated data sets is more complex...

@@

>>>>

Furthermore, the proportion of the environmental effect in the column space of $\boldsymbol{P}_{\bX}$ (i.e., the portion that will cause confounding) is given by
\begin{equation}
    \label{eqn:ratio}
    \text{CP} = ||\bX \boldsymbol{\tau}||_2^2 / ||\bZ \bgamma||_2^2,
\end{equation}
which we refer to as the Confounding Proportion (CP). The CP can range from 0 (no confounding) to 1 (entirely confounded). In the extreme case where $\bZ \bgamma$ is completely independent of population structure, $\bZ \bgamma \in \mathcal{N}(\boldsymbol{P}_{\bX})$ and $\bX \boldsymbol{\tau} = \boldsymbol{0}$ so that $\mathbb{E}(\bbetaHat^*) = \bbeta$. In this scenario, the SNP-phenotype associations will be unbiased, but the residual error will be inflated.

@@

 To evaluate the performance of penalized LMM methods in the presence of these different forms of confounding complexity, we consider environmental effects which are concordant or discordant in nature in the presence of both exponential and dichotomous effect sizes, for a total of four distinct $\bgamma$ patterns. Figure \ref{fig:gamma_structures} provides a visual summary of the four coefficient patterns, where the subpopulations on the x-axis are ordered by genetic similarity so that subpopulation 1 is most genetically similar to subpopulation 2, and least similar to subpopulation 10. This may also be thought of as subpopulations ordered spatially along a 1-dimensional geographic region.
\begin{figure}[H]
    \centering
    \includegraphics[scale = 0.9]{figures/gamma_structure.png}
    \caption{An illustration of four distinct patterns of environmental effects and corresponding $\bgamma$ values in the presence of 10 subpopulations ordered by genetic similarity.}
    \label{fig:gamma_structures}
\end{figure}
\noindent In the case of concordant and dichotomous effects, Figure \ref{fig:gamma_structures} shows that the first five ordered subpopulations are subject to an effect size of 1, while the second five are subject to an effect size of $-$1. In the case of discordant dichotomous effects, the first five subpopulations experience an effect size with an absolute value of \anna{1}, and the second five an absolute effect size of \anna{0.5}. However, the directions of these absolute effects sizes alternate among the ordered subpopulations so that genetically similar subpopulations do not experience similar environmental effects. \anna{not quite 1 and 0.5 because of centering and standardization...how to word this so it's not confusing? - trying better version of dichotomous-discordant...}

@@

Due to the differing sample sizes and subpopulation structures, a direct comparison of the Empirical CP with those from the simulated data sets is more complex... this increases with equal subpopulation sizes

@@

 \anna{correlation between columns of X..., mspe vs mse- for various reasons CP is a reasonable measure but not perfect...}
